Article Title

### 2013-12-12 plos.tex
updtated introduction

\section*{Introduction}
Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are the most familiar descriptive statistics of neural population activity \cite{Averbeck:2006,Zohary:1994,Kohn:2005,Bair:2001,Ecker:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of stimulus response variability between pairs of neurons, have been shown to have profound theoretical implications for stimulus coding \cite{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011,Ecker:2011}. In addition, neural correlations are hypothesized to reflect key features of functional connectivity in neural circuits.  Such interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distance separating the neurons \cite{Smith:2008,Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus tuning similarity \cite{Kohn:2005,Ko:2011}, cortical layer specificity \cite{Hansen:2012,Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009}, changes due to sensory stimulation and global brain states \cite{Goard:2009,Kohn:2009,Ecker:2010,Renart:2010}, and others.

@@

In this study, we focused on low-dimensional representations composed of two types of elements: the \emph{low-rank correlation structure} and \emph{partial correlations} between pairs of neurons. The former is most suitable for expressing common circuit fluctuations whereas the latter are most suitable for expressing detailed interactions between specific neurons. Low-rank neural correlation structure extracted by principal component analysis or factor analysis has been analyzed for temporal dynamics in in several studies \cite{Nicolelis:1995,Yu:2006,Yu:2009}.

Partial correlations between pairs of neurons are linear correlations remaining after subtracting the linear effect of all the other neurons. In systems where effects exerted by variables on one another have strong linear components, partial correlations describe conditional dependency: zero  partial correlation between two variables implies conditional independence.  Conditional independence between two variables, if all relevant variables are observed, implies a lack of direct interaction between the underlying processes. Therefore, estimation of networks of partial correlations (sometimes called \emph{association networks}) have been applied to infer interactions between components of complex systems such as gene interaction networks \cite{Schafer:2005,Peng:2009} and functional connectivity between brain regions in fMRI signals \cite{Varoquaux:2012,Ryali:2012}. In multivariate normal distributions\footnote{or any other \emph{elliptical} distribution of the form $g(x-\mu)^\T\Sigma^{-1}(x-\mu))$, where $g$ is an arbitrary function}, are completely defined by linear effects. A multivariate normal distribution defined through a sparse network of non-zero partial correlations is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With other distributions where nonlinear effects are more significant, the correspondence between conditional dependence and partial correlations diminishes or breaks down \cite{Loh:2012}.

For example, factor analysis or principal component analysis may reveal common activity fluctuations across the entire population \cite{Yu:2009}.  The correlation matrix also allows computing the \emph{partial correlations} between specific pairs of neurons, \emph{i.e.}\;correlations that remain after accounting for the linear effects of all other recorded cells.  Under several assumptions (predominance of linear effects (sufficiently complete sampling of interacting neurons, and sufficient sample size), partial correlations may better reflect the direct statistical associations or interactions between pairs of neurons. In multivariate normal distributions, zero partial correlations between pairs of variables indicate conditional independence or lack of interaction between the pair and the network of non-zero interactions constitute Gaussian Graphical Models or Gauss-Markov Random Fields \cite{Koller:2009}. With other distributions, the correspondence between conditional dependence and partial correlations diminishes or breaks down \cite{Loh:2012}. Estimation of networks of partial correlations (sometimes called \emph{association networks}) have been applied to map gene interactions \cite{Schafer:2005,Peng:2009} or to establish functional connectivity between brain regions from fMRI data \cite{Varoquaux:2012,Ryali:2012}.

Multineuronal recordings also present major challenges for the estimation and interpretation of correlations.  As the population size increases, the amount of recorded data grows only linearly whereas the number of estimated coefficients grows quadratically.  This mismatch leads to increased opportunities for spurious correlations, overestimation of shared activity (\emph{i.e.}\;overestimation of large eigenvalues) \cite{Ledoit:2004}, and poorly conditioned estimates of the partial correlations \cite{Schafer:2005}.

@@

@@

Estimation can be improved through \emph{regularization}: the deliberate biasing of the estimate toward one of several possible low-dimensional approximations referred to here as \emph{target estimates} \cite{Schafer:2005,Bickel:2006}.  The usual estimate, the \emph{sample covariance matrix} has the advantage of being unbiased but, on average, it falls far from the true covariance matrix due to its sensitivity to sampling noise in the data.  Low-dimensional estimates of various forms are typically less susceptible to sampling noise but are also liable to introduce their respective biases away from the true covariance matrix.  Regularization works by striking a favorable balance between bias and variability. Regularization can produce \emph{some} improvement even with an arbitrary target estimate that has no relation to the true covariance structure in the data.  Yet, when the target estimate is well suited for capturing the important features of the true covariance matrix with few terms, it will introduce minimal bias and outperform other estimators.

To illustrate key concepts, Figure \ref{fig:01} illustrates a regularization scheme based on \emph{covariance selection} \cite{Dempster:1972}. In covariance selection, the estimate is produced by fitting only an optimal subset of the coefficients of the \emph{precision matrix} (the inverse of the covariance matrix) while setting the remaining coefficients to zero. Scaling the precision matrix so that its diagonal is composed of all $-1$'s produces the matrix of partial correlations. Thus zeros in the precision matrix produce zero partial correlations between the corresponding pairs of neurons.  Panel \ref{fig:01}\,A depicts the sample correlation matrix of a period of somatic calcium signals of a group of 298 neurons in mouse visual cortex.  Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are much higher than typical firing rate correlations on shorter temporal scales. Panel \ref{fig:01}\,B magnifies the fragment of the sample correlation matrix outlined in \ref{fig:01}\,A whereas \ref{fig:01}\,C shows the corresponding fragment of the regularized estimate. The regularized estimate is produced from the same data by zeroing the optimal set of 31,501 (71.2\%) off-diagonal coefficients of the precision matrix and fitting the remaining coefficients. Panels \ref{fig:01}\,D and E show the same fragments of the respective estimates of the partial correlation matrix. The apparent similarity of the two estimates of the correlation matrix belies the utterly different partial correlation structure. The regularized estimate is less sensitive to sample noise in the data, is likely to be closer to truth in most cases than the unbiased estimate, which may be ascertained by cross-validation studies.

It may be tempting to interpret the regularized sparse estimate of partial correlations as a more accurate representation of the functional connectivity than the original correlation coefficients: after all, the regularized solution consistently produces better estimates and has fewer parameters which are easier to interpret. However, an alternative regularization scheme using 43 latent factors interacting linearly with all neurons while having about the same number of parameters, performed even better than covariance selection on this dataset.

@@

\subsection*{Covariance estimation}
We aim to estimate the true covariance matrix 
\begin{equation}\label{eq:true-covariance}
\Sigma = \E{(x-\mu)(x-\mu)^\T}
\end{equation}
where $\E{\cdot}$ denotes expectation; $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates in bins of duration $\Delta t$. The vector of mean firing rates is $\mu = \E{x}$.

The usual estimator of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf 0}$ computed from the empirical sample of observations $x(1),\ldots,x(n)$ as
\begin{equation}
C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\mu)(x(t)-\mu)^\T 
\end{equation}
where $\nu$ is the number of degrees of freedom per neuron in the sample ($\nu=n-1$ if observations are independent).  
The sample covariance matrix is unbiased, \emph{i.e.} $\E{C_{\sf 0}}=\Sigma$.
For finite sample sizes, however, $C_{\sf 0}$ is not as close to $\Sigma$ as a number of biased estimators.

We considered four regularized estimators based on distinct families of low-dimensional target estimates: `independent', `latent factors', `sparse interactions', and `sparse+latent' (Fig.~\ref{fig:02}\;row 1).

>>>>

\section*{Introduction}
Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are the most familiar descriptive statistics of neural population activity \cite{Averbeck:2006,Zohary:1994,Kohn:2005,Bair:2001,Ecker:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of stimulus response variability between pairs of neurons, have been shown to have profound theoretical implications for stimulus coding \cite{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011,Ecker:2011}. In addition, neural correlations are hypothesized to reflect key features of functional connectivity in neural circuits.  Such interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008,Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus tuning similarity \cite{Kohn:2005,Ko:2011}, cortical layer specificity \cite{Hansen:2012,Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009}, changes due to sensory stimulation and global brain states \cite{Goard:2009,Kohn:2009,Ecker:2010,Renart:2010}, and others.

@@

For example, one form of low-dimensional representation of the correlation matrix is its low-rank approximation. Low-rank approximations are particularly suitable for capturing shared fluctuations across the recorded population. Extracted through principal component or factor analysis, low-rank components of neural correlations have been analyzed in a number of studies with particular attention paid to their temporal dynamics \cite{Yu:2006,Yu:2009}.

Alternatively, the correlation matrix can be formulated through the partial correlations between pairs of neurons. The partial correlation between a pair of neurons is the linear correlation remaining after accounting for correlations to all the other neurons. Partial correlations carry particular significance when conditional dependencies between all the variables are thought to be dominated by linear effects. In such systems, zero  partial correlation between two variables implies their conditional independence.  Conditional independence between a pair of variables suggests a lack of direct interaction between the underlying processes. Therefore, estimation of networks of partial correlations (sometimes called \emph{association networks}) have been used to infer interactions between components in complex systems such as gene interaction networks \cite{Schafer:2005,Peng:2009} and functional connectivity between brain regions in fMRI signals \cite{Varoquaux:2012,Ryali:2012}. In \emph{elliptical distributions} (of the form $g\left((x-\mu)^\T\Sigma^{-1}(x-\mu)\right)$, where $g$ is an arbitrary function $\Sigma$ is the covariance matrix) and the multivariate normal distribution in particular, interactions are completely defined by linear effects and partial interactions express conditional dependencies. A multivariate normal distribution specified through its graph of non-zero partial correlations is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With increasing departure from linearity, the correspondence between conditional dependence and partial correlations diminishes or completely breaks down \cite{Loh:2012}. For example, the coupling terms in pairwise Ising models have no direct relationship to the partial correlations.

@@

The true correlation matrix is the normalized version of the true covariance matrix defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates in bins of duration $\Delta t$; and $\mu$ is the vector of mean firing rates.  The usual estimator of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf 0}$ computed from the empirical sample of observations $x(t),\; t=n(1,\ldots,n)$ as
\begin{equation}
    C_{\sf 0} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom per neuron in the sample ($\nu=n-1$ if observations are independent). In this study, we keep the sample mean $\bar x$ as the estimate of the true mean $\mu$, but seek a better estimate of the covariance matrix than $C_{\sf 0}$.

Estimation of the covariance matrix from a large population presents a number of numerical challenges.  The amount of recorded data grows only linearly with the population size whereas the number of estimated coefficients increases quadratically.  This mismatch leads to increased opportunities for spurious correlations, to overestimation of shared activity (\emph{i.e.}\;overestimation of large eigenvalues) \cite{Ledoit:2004}, and to poorly conditioned estimates of the partial correlations \cite{Schafer:2005}.

Estimation can be improved through \emph{regularization}: the deliberate biasing of the estimate toward a low-dimensional approximation referred to here as the \emph{target estimate} \cite{Schafer:2005,Bickel:2006}.  The usual estimate $C_{\sf 0}$ has the advantage of being unbiased, \emph{i.e.}\;$\E{C_0}=\Sigma$ but, on average, it falls far from $\Sigma$ due to its sensitivity to sampling noise in the data.  Low-dimensional estimates of various forms are typically less susceptible to sampling noise but are also liable to introduce their respective biases away from truth $\Sigma$.  Regularization works by striking a favorable balance between bias and variability. Regularization can produce \emph{some} improvement even with an arbitrary target estimate that has no relation to the true dependencies in the data.  Yet, when the target estimate is well suited for capturing the important features of the true covariance matrix with few terms, it will introduce minimal bias and outperform other estimators.

Figure \ref{fig:01} illustrates the challenges of estimating the correlation matrix from a finite sample. Panel \ref{fig:01}\,A depicts the sample correlation matrix $R_{\sf 0}$ of a period of somatic calcium signals of a group of 298 neurons in mouse visual cortex. $R_{\sf 0}$ is computed by normalizing the covariance matrix $C_{\sf 0}$ by the sample variances:
\begin{equation}
    R_{\sf 0} = \left(I\circ C_{\sf 0}\right)^{-\frac 1 2} C_{\sf 0} \left(I\circ C_{\sf 0}\right)^{-\frac 1 2}
\end{equation}
Here $\circ$ denotes entrywise matrix product (Hadamard product). Thus $I\circ C_{\sf 0}$ represents the diagonal matrix with the diagonal elements from $C_{\sf 0}$, \emph{i.e.} the sample variances. 
Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are much higher than typical firing rate correlations on shorter temporal scales. To show greater detail, panel \ref{fig:01}\,B magnifies the fragment of the sample correlation matrix that is outlined in \ref{fig:01}\,A. The matrix of partial correlations $P$ is obtained by scaling the inverse of the correlation matrix:
\begin{equation}
    P = -D^{-1/2} R^{-1} D^{-1/2}
\end{equation}
where $D=I\circ R^{-1}$ is the diagonal matrix with the diagonal elements of the inverse of the correlation matrix $R$. Thus $P$ contains $-1$s on its diagonal and the off-diagonal elements contain the pairwise partial correlations. Figure \ref{fig:01}\,C shows the magnified fragment of the partial correlation matrix $P$.

Panels \ref{fig:01}\,D and E depict a different estimate obtained from the same data by \emph{covariance selection} \cite{Dempster:1972}. In covariance selection, only a subset of partial correlations is fitted while the rest are fixed at zero. The resulting estimate closely reproduces the sample correlation matrix (Fig.~\ref{fig:01}\,D), but its partial correlation structure appears substantially different(Fig.~\ref{fig:01}\,E). In this particular example  31,501 of the 44,253 (71.2\%) total partial correlations are set to zero. This alternative estimate is attractive for a few reasons. First, it has fewer free parameters and is therefore less susceptible to sampling noise. Additionally, assuming that partial correlations do indeed reflect elements of the functional connectivity, the second estimate also appears more physiologically plausible: in neural circuits, each neuron only interacts with a fraction of the entire population.

@@

@@

\subsection*{Covariance estimation}
We considered four regularized estimators based on distinct families of low-dimensional target estimates: `independent', `latent factors', `sparse interactions', and `sparse+latent' (Fig.~\ref{fig:02}\;row 1).
